Visualisation of Brain Statistics with R-package `ggseg`

### 2019-04-15 draft_1.tex
moved some files

%

@@

Neuroscientific analysis usually requires the use of multiple tools to analyse, visualise, and summarise data.
This often makes the process of preparing results for publication laborious, as results must be exported and imported in various formats.
In neuroimaging, images of probands' brains are collected and merged together to provide three dimensional representations.
Much of the neuroimaging analyses are not necessarily done on the individual voxel level (3-dimensional pixels), but rather on pre-defined brain segmentations, called brain atlases.
These brain atlases are plentiful and represent different ways of segmenting the brain into functionally or structurally similar regions (REF).
The use of these is wide-spread, as these atlases provide larger meaningful divisions of the brain (REF).
While neuroimaging analyses on the voxel-level are usually computed by special software for such analyses, analyses of brain atlas data is usually done in standard statistical software, like R (REF), python (REF), or Matlab (REF).

@@

There are several tools that aid R-users in plotting neuroimaging data directly through R using the grammar of graphics as implemented in ggplot2 (REF), such as ggBrain(REF) and ggneuro(REF)(see \href{https://neuroconductor.org/}{neuroconductor} for compiled neuroimaging packages for R).
These are based on plotting imaging files, not results from analyses of brain atlases.
Here we introduce the ggseg-package for visualizing results from brain atlas analyses.
The ggseg-package was developed to create templates that others might use to project their brain atlas results on to.
While its plotting functions are what the users are drawn to, it is the pre-compiled number of data sets for different brain atlases that provides the real functionality needed for this visualization.
The package also includes functions to plot atlases in tri-surface mesh plots using plotly (REF).

@@

There is a multitude of brain atlases to chose from, all with different purposes and segmented based on different criteria.
We have initially focused on commonly used and well-established atlases with manageable numbers of segmentations.
The ggseg-package thus features data for two well-established parcellations, as implemented in the neuroimaging software Freesurfer (REF); dkt and aseg.
The Desikan-Killany cortical atlas (dkt) (REF) and the Automatic Segmentation of Subcortical Structures (aseg) (REF) together cover all grey matter of the brain, and are commonly used in analyses elucidating the associations between thickness, volume, or area of these segments and mental health and cognitive function.

@@

Creating plots of the brain as 2-dimensional can at times make it difficult to properly understand where an atlas segment is located in the true 3-dimensional brain.
This is particularly true when looking within the brain as subcortical structures or white matter tracts.
To provide the option of viewing and printing the plots in 3-dimensional space, another function was created using \href{https://plotly-book.cpsievert.me/}{plotly}.
The data is somewhat more complex than the 2D ggplot polygon version \texttt{ggseg}, as it includes options for brain inflation, glass brains and much more, to ease viewing and understanding.
Another positive effect of basing this function on plotly, is the interactivity of the plot, which also makes it nice for public dissemination.
A lot of credit goes to A.M.Winkler and his \href{https://brainder.org/}{Brainder} work, which supplied us with the first examples of going from Freesurfer \texttt{.srf}-files to \texttt{.ply}-files, and whose scripts massively aided us in making this work.

The function \texttt{ggseg3d()}, is based in the plotly, it is recommended to get a little familiarized with with \href{https://plotly-book.cpsievert.me/}{plotly}.

@@

There are many plotly options that may help you adapt the plot to look the way you wish.
In addition to Carson Sievert's \href{https://plotly-book.cpsievert.me/}{book on plotly in R}, we recommend resources for \href{https://plot.ly/r/axes/\#modifying-axes-for-3D-plots}{modifying axes in 3D plots}, the \href{https://plot.ly/r/trisurf/}{basic introduction to tri-surface plots}, and this tutorial on \href{https://moderndata.plot.ly/trisurf-plots-in-r-using-plotly/}{tri-surface plots with plotly in R}.

@@

>>>>

% 
\usepackage{natbib}
\bibliographystyle{plainnat}

@@

Neuroscientific analysis usually requires the use of multiple tools to analyse, visualise, and summarise data.
This often makes the process of preparing results for publication laborious, as results must be exported and imported in various formats.
In neuroimaging, images of probands' brains are collected and merged together to provide three dimensional representations.
Much of the neuroimaging analyses are not necessarily done on the individual voxel level (3-dimensional pixels), but rather on pre-defined brain segmentations, called brain atlases.
These brain atlases are plentiful and represent different ways of segmenting the brain into functionally or structurally similar regions (\citet{dkt}).
The use of these is wide-spread, as these atlases provide larger meaningful divisions of the brain.
While neuroimaging analyses on the voxel-level are usually computed by special software for such analyses, analyses of brain atlas data is usually done in standard statistical software, like R (\citet{R}), python (\citet{python}), or MATLAB (\citet{MATLAB}).

@@

There are several tools that aid R-users in plotting neuroimaging data directly through R using the grammar of graphics as implemented in ggplot2 (\citet{ggplot}), such as ggBrain (\citet{ggBrain}) and ggneuro (\citet{ggneuro}; see neuroconductor \citeyearpar{neuroconductor} for curated neuroimaging packages for R).
These are based on plotting imaging files, not results from analyses of brain atlases.
Here we introduce the ggseg-package for visualizing results from brain atlas analyses.
The ggseg-package was developed to create templates that others might use to project their brain atlas results on to.
While its plotting functions are what the users are drawn to, it is the pre-compiled number of data sets for different brain atlases that provides the real functionality needed for this visualization.
The package also includes functions to plot atlases in tri-surface mesh plots using plotly (\citet{plotly}).

@@

There is a multitude of brain atlases to chose from, all with different purposes and segmented based on different criteria.
We have initially focused on commonly used and well-established atlases with manageable numbers of segmentations.
The ggseg-package thus features data for two well-established parcellations, as implemented in the neuroimaging software Freesurfer \citep{fischl:99}; dkt and aseg.
The Desikan-Killany cortical atlas (dkt; \citet{dkt}) and the Automatic Segmentation of Subcortical Structures (aseg; \citet{aseg}) together cover all grey matter of the brain, and are commonly used in analyses elucidating the associations between thickness, volume, or area of these segments and mental health and cognitive function.

@@

Creating plots of the brain as 2-dimensional can at times make it difficult to properly understand where an atlas segment is located in the true 3-dimensional brain.
This is particularly true when looking within the brain as subcortical structures or white matter tracts.
To provide the option of viewing and printing the plots in 3-dimensional space, another function was created using plotly \citep{plotly}.
The data is somewhat more complex than the 2D ggplot polygon version \texttt{ggseg}, as it includes options for brain inflation, glass brains and much more, to ease viewing and understanding.
Another positive effect of basing this function on plotly, is the interactivity of the plot, which also makes it nice for public dissemination.
A lot of credit goes to A.M.Winkler and his Brainder (\citet{brainder}) work, which supplied us with the first examples of going from Freesurfer \texttt{.srf}-files to \texttt{.ply}-files, and whose scripts massively aided us in making this work.

The function \texttt{ggseg3d()}, is based in the plotly, it is recommended to get a little familiarized with with plotly (\citet{plotly}).

@@

There are many plotly options that may help you adapt the plot to look the way you wish.
In addition to Carson Sievert's book on plotly in R (\citeyearpar{plotly}) we recommend resources for modifying axes in 3D plots (\citet{plotly-ax}), the basic introduction to tri-surface plots (\citet{plotly-tri}), and this tutorial on tri-surface plots with plotly in R (\citet{plotly-trisurf}).

@@

\hypertarget{author-contributions}{%
\section{Author Contributions}\label{author-contributions}}

Didac Vidal-Piñeiro generated the idea for the plotting tool, and the initial scripts that made the plots.
He has also been responsible for converting images from neuroimaging software into the type of data necessary for plotting polygons and mesh plots in R.
Athanasia M. Mowinckel adapted the initial scripts and made the functions into package format and making the functions more user-friendly.
She also was responsible for adding the mesh-plot functionality through plotly, and developing the pipeline for making that possible.
A. M. Mowinckel wrote the first draft of the paper, and both have since critically edited it.

\hypertarget{conflicts-of-interest}{%
\section{Conflicts of Interest}\label{conflicts-of-interest}}

The authors declare that there were no conflicts of interest with respect to the authorship or the publication of this article.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

We also thanks John Muschelli for his initial improvements in the package code and preparing it for incorporation in neuroconductor (\citeyearpar{neuroconductor}), and Richard Beare for submitting the first community created atlas, and providing helpful comments to the base code.

\hypertarget{funding}{%
\section{Funding}\label{funding}}

This work is funded by EU Horizon 2020 Grant `Healthy minds 0-100 years: Optimising the use of European brain imaging cohorts (Lifebrain)', with grant agreement number 732592.

\hypertarget{prior-versions}{%
\section{Prior versions}\label{prior-versions}}

Some of the content presented here also apprears in the package vignette of \texttt{ggseg}, which may be accessed through R or in the package website (\citet{ggseg}).
Athanasia Monika Mowinckel also has several tutorials on her blog regarding ggseg creationg and functionality (\citet{ggsegAnim}, \citet{ggsegIntro}).

\renewcommand\refname{References}
\bibliography{references.bib}

### 2019-04-15 draft_1_AMM.rmd
moved some files

---
title: Visualisation of Brain Statistics with R-package `ggseg`
preprint: true
author: 
  - name: Athanasia M. Mowinckel
    affiliation: 1
    corresponding: true
    email: a.m.mowinckel@psykologi.uio.no
  - name: Didac Vidal Piñeiro
    affiliation: 1
affiliation:
  - code: 1
    address: Center for Lifespan Changes in Brain and Cognition, Univeristy of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  The abstract of the article.
  It can also be on *multiple* lines.
header-includes: 
- \usepackage{lipsum} 
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
bibliography: sample.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    fig_caption: true
  rticles::peerj_article: default
---
```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE)
library(kableExtra)
library(tidyverse)
library(plotly) 
library(ggplot2)
library(knitr)
options(tinytex.verbose = TRUE)

@@

Neuroscientific analysis usually requires the use of multiple tools to analyse, visualise, and summarise data. 
This often makes the process of preparing results for publication laborious, as results must be exported and imported in various formats. 
In neuroimaging, images of probands' brains are collected and merged together to provide three dimensional representations.
Much of the neuroimaging analyses are not necessarily done on the individual voxel level (3-dimensional pixels), but rather on pre-defined brain segmentations, called brain atlases.
These brain atlases are plentiful and represent different ways of segmenting the brain into functionally or structurally similar regions (REF). 
The use of these is wide-spread, as these atlases provide larger meaningful divisions of the brain (REF). 
While neuroimaging analyses on the voxel-level are usually computed by special software for such analyses, analyses of brain atlas data is usually done in standard statistical software, like R (REF), python (REF), or Matlab (REF).

@@

There are several tools that aid R-users in plotting neuroimaging data directly through R using the grammar of graphics as implemented in ggplot2 (REF), such as ggBrain(REF) and ggneuro(REF)(see [neuroconductor](https://neuroconductor.org/) for compiled neuroimaging packages for R). 
These are based on plotting imaging files, not results from analyses of brain atlases. 
Here we introduce the ggseg-package for visualizing results from brain atlas analyses.
The ggseg-package was developed to create templates that others might use to project their brain atlas results on to. 
While its plotting functions are what the users are drawn to, it is the pre-compiled number of data sets for different brain atlases that provides the real functionality needed for this visualization.
The package also includes functions to plot atlases in tri-surface mesh plots using plotly (REF).

## Brain atlas selection
There is a multitude of brain atlases to chose from, all with different purposes and segmented based on different criteria. 
We have initially focused on commonly used and well-established atlases with manageable numbers of segmentations. 
The ggseg-package thus features data for two well-established parcellations, as implemented in the neuroimaging software Freesurfer (REF); dkt and aseg. 
The Desikan-Killany cortical atlas (dkt) (REF) and the Automatic Segmentation of Subcortical Structures (aseg) (REF) together cover all grey matter of the brain, and are commonly used in analyses elucidating the associations between thickness, volume, or area of these segments and mental health and cognitive function.

@@

Creating plots of the brain as 2-dimensional can at times make it difficult to properly understand where an atlas segment is located in the true 3-dimensional brain. 
This is particularly true when looking within the brain as subcortical structures or white matter tracts.
To provide the option of viewing and printing the plots in 3-dimensional space, another function was created using [plotly](https://plotly-book.cpsievert.me/).
The data is somewhat more complex than the 2D ggplot polygon version `ggseg`, as it includes options for brain inflation, glass brains and much more, to ease viewing and understanding.
Another positive effect of basing this function on plotly, is the interactivity of the plot, which also makes it nice for public dissemination.
A lot of credit goes to A.M.Winkler and his [Brainder](https://brainder.org/) work, which supplied us with the first examples of going from Freesurfer `.srf`-files to `.ply`-files, and whose scripts massively aided us in making this work.

The function `ggseg3d()`, is based in the plotly, it is recommended to get a little familiarized with with [plotly](https://plotly-book.cpsievert.me/).

@@

There are many plotly options that may help you adapt the plot to look the way you wish. 
In addition to Carson Sievert's [book on plotly in R](https://plotly-book.cpsievert.me/), we recommend resources for [modifying axes in 3D plots](https://plot.ly/r/axes/#modifying-axes-for-3D-plots), the [basic introduction to tri-surface plots](https://plot.ly/r/trisurf/), and this tutorial on [tri-surface plots with plotly in R](https://moderndata.plot.ly/trisurf-plots-in-r-using-plotly/).

@@

>>>>

---
title: Visualisation of Brain Statistics with R-package `ggseg`
preprint: true
author: 
  - name: Athanasia M. Mowinckel
    affiliation: 1
    corresponding: true
    email: a.m.mowinckel@psykologi.uio.no
  - name: Didac Vidal Piñeiro
    affiliation: 1
affiliation:
  - code: 1
    address: Center for Lifespan Changes in Brain and Cognition, Univeristy of Oslo, PO. box 1094 Blindern, 0317 Oslo, Norway
abstract: >
  The abstract of the article.
  It can also be on *multiple* lines.
header-includes: 
- \usepackage{lipsum} 
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
bibliography: references.bib
output:
  bookdown::pdf_book:
    base_format: rticles::peerj_article # for using bookdown features like \@ref()
    citation_package: natbib
    fig_caption: true
  rticles::peerj_article: default
---
```{r setup, include=F}
knitr::opts_chunk$set(eval=TRUE, echo=TRUE)
library(kableExtra)
library(tidyverse)
library(plotly) 
library(ggplot2)
library(knitr)
options(tinytex.verbose = TRUE)

@@

Neuroscientific analysis usually requires the use of multiple tools to analyse, visualise, and summarise data. 
This often makes the process of preparing results for publication laborious, as results must be exported and imported in various formats. 
In neuroimaging, images of probands' brains are collected and merged together to provide three dimensional representations.
Much of the neuroimaging analyses are not necessarily done on the individual voxel level (3-dimensional pixels), but rather on pre-defined brain segmentations, called brain atlases.
These brain atlases are plentiful and represent different ways of segmenting the brain into functionally or structurally similar regions (@dkt). 
The use of these is wide-spread, as these atlases provide larger meaningful divisions of the brain. 
While neuroimaging analyses on the voxel-level are usually computed by special software for such analyses, analyses of brain atlas data is usually done in standard statistical software, like R (@R), python (@python), or MATLAB (@MATLAB).

@@

There are several tools that aid R-users in plotting neuroimaging data directly through R using the grammar of graphics as implemented in ggplot2 (@ggplot), such as ggBrain (@ggBrain) and ggneuro (@ggneuro; see neuroconductor -@neuroconductor for curated neuroimaging packages for R). 
These are based on plotting imaging files, not results from analyses of brain atlases. 
Here we introduce the ggseg-package for visualizing results from brain atlas analyses.
The ggseg-package was developed to create templates that others might use to project their brain atlas results on to. 
While its plotting functions are what the users are drawn to, it is the pre-compiled number of data sets for different brain atlases that provides the real functionality needed for this visualization.
The package also includes functions to plot atlases in tri-surface mesh plots using plotly (@plotly).

## Brain atlas selection
There are multitudes of brain atlases to chose from, all with different purposes and segmented based on different criteria. 
We have initially focused on commonly used and well-established atlases with manageable numbers of segmentations. 
The ggseg-package thus features data for two well-established parcellations, as implemented in the neuroimaging software Freesurfer [@fischl:99]; dkt and aseg. 
The Desikan-Killany cortical atlas (dkt; @dkt) and the Automatic Segmentation of Subcortical Structures (aseg; @aseg) together cover all grey matter of the brain, and are commonly used in analyses elucidating the associations between thickness, volume, or area of these segments and mental health and cognitive function.

@@

Creating plots of the brain as 2-dimensional can at times make it difficult to properly understand where an atlas segment is located in the true 3-dimensional brain. 
This is particularly true when looking within the brain as subcortical structures or white matter tracts.
To provide the option of viewing and printing the plots in 3-dimensional space, another function was created using plotly [@plotly].
The data is somewhat more complex than the 2D ggplot polygon version `ggseg`, as it includes options for brain inflation, glass brains and much more, to ease viewing and understanding.
Another positive effect of basing this function on plotly, is the interactivity of the plot, which also makes it nice for public dissemination.
A lot of credit goes to A.M.Winkler and his Brainder (@brainder) work, which supplied us with the first examples of going from Freesurfer `.srf`-files to `.ply`-files, and whose scripts massively aided us in making this work.

The function `ggseg3d()`, is based in the plotly, it is recommended to get a little familiarized with with plotly (@plotly).

@@

There are many plotly options that may help you adapt the plot to look the way you wish. 
In addition to Carson Sievert's book on plotly in R (-@plotly) we recommend resources for modifying axes in 3D plots (@plotly-ax), the basic introduction to tri-surface plots (@plotly-tri), and this tutorial on tri-surface plots with plotly in R (@plotly-trisurf).

@@

# Author Contributions
Didac Vidal-Piñeiro generated the idea for the plotting tool, and the initial scripts that made the plots. 
He has also been responsible for converting images from neuroimaging software into the type of data necessary for plotting polygons and mesh plots in R. 
Athanasia M. Mowinckel adapted the initial scripts and made the functions into package format and making the functions more user-friendly.
She also was responsible for adding the mesh-plot functionality through plotly, and developing the pipeline for making that possible. 
A. M. Mowinckel wrote the first draft of the paper, and both have since critically edited it.

# Conflicts of Interest
The authors declare that there were no conflicts of interest with respect to the authorship or the publication of this article.

# Acknowledgements
We also thanks John Muschelli for his initial improvements in the package code and preparing it for incorporation in neuroconductor (-@neuroconductor), and Richard Beare for submitting the first community created atlas, and providing helpful comments to the base code.

# Funding 
This work is funded by EU Horizon 2020 Grant 'Healthy minds 0-100 years: Optimising the use of European brain imaging cohorts (Lifebrain)', with grant agreement number 732592.

# Prior versions
Some of the content presented here also apprears in the package vignette of `ggseg`, which may be accessed through R or in the package website (@ggseg). 
Athanasia Monika Mowinckel also has several tutorials on her blog regarding ggseg creationg and functionality (@ggsegAnim, @ggsegIntro).

# References
